Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-07 paper/anglican.tex
updated the colors

		defmulti,defn,defn-,defonce,defquery,defprotocol,defrecord,defstruct,deftype,delay,delay?,deliver,deref,derive,%

@@

		make-array,make-hierarchy,map,map?,mapcat,max,max-key,memfn,memoize,merge,merge-with,meta,%

@@

		replace,replicate,require,reset!,reset-meta!,resolve,rest,resultset-seq,reverse,reversible?,%

@@

		split-with,str,stream?,string?,struct,struct-map,subs,subseq,subvec,supers,swap!,symbol,symbol?,%

>>>>

		defmulti,defn,defn-,defonce,defquery,defproc,defprotocol,defrecord,defstruct,deftype,delay,delay?,deliver,deref,derive,%

@@

		make-array,make-hierarchy,map,map?,mapcat,max,max-key,mem,memfn,memoize,merge,merge-with,meta,%

@@

		replace,replicate,require,reset!,reset-meta!,resolve,rest,resultset-seq,retrieve,reverse,reversible?,%

@@

		split-with,store,str,stream?,string?,struct,struct-map,subs,subseq,subvec,supers,swap!,symbol,symbol?,%

### 2017-02-07 paper/paper.tex
updated the colors

\documentclass[preprint]{sigplanconf}
\linespread{1.00}

@@

\usepackage{listings}
\usepackage{lmodern}
\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily, columns=flexible, showstringspaces=false}

@@

An Anglican program, or \textit{query}, is compiled into a
Clojure function. When inference is performed with a provided
algorithm, this produces a sequence of samples. Anglican shares
a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican
can look like the following code:
%\begin{minipage}{\linewidth}
\begin{lstlisting}[style=default]
(defquery model data
  "chooses a distribution
   which describes the data"
  (let [;;; Guess a distribution.
        dist (sample (categorical
                       [[normal 0.5]
                        [gamma 0.5]]))
        a (sample (gamma 1 1))
        b (sample (gamma 1 1))
        d (dist a b)]

@@

Anglican is a language for expressing probabilistic programs.
A probabilistic program is a stateful \textit{deterministic} computation
$\mathcal{P}$ with the following properties~\cite{TMP+15}:
\begin{itemize}
\item Initially, $\mathcal{P}$ expects no arguments.
\item On every call, $\mathcal{P}$ returns either a distribution
  and an address $(F,\alpha)$, a distribution and a value $(G, y)$, a
  value $z$, or $\bot$.
\item Upon returning $F$, $\mathcal{P}$ expects a value $x$ drawn from $F$ as
  the argument to the next call.
\item Upon returning $(G, y)$ or $z$, $\mathcal{P}$ is invoked again
  without arguments.
\item Upon returning $\bot$, $\mathcal{P}$ terminates.
\end{itemize}
A program is run by calling $\mathcal{P}$ repeatedly until
termination.

A program need not generate the same sequence of random variables in every
execution. For this reason we assume that each random variable $x$ is assigned
a unique label $\alpha$, which we call an address, that induces a
correspondence between variables in different executions. Every execution
implicitly produces a sequence of triples of distributions, values of
\textit{latent} random variables, and addresses, $(F_i, x_i, \alpha_i)$.  We
call this sequence a \textit{trace} and denote it by $\pmb{x}$. A trace induces
a sequence of pairs $(G_j, y_j)$ of distributions and values of
\textit{observed} random variables. We call this sequence an \textit{image} and
denote it by $\pmb{y}$. For notational simplicity we assume that the program
always generates the same ordered set $\pmb{z}$ of \textit{output} values
$z_k$.

@@

The Anglican language is a subset of Clojure\footnote{It would be
possible to support almost full Clojure by expanding all macros
in the Anglican source code. However, in Clojure, unlike in
Scheme~\cite{SDF+10} or Common Lisp~\cite{PC94}, the result
of macro-expansion of derived special forms is not well
specified and implementation specific.}.  Anglican queries
are defined within \texttt{defquery} as shown in the previous section,
using \texttt{if}, \texttt{when}, \texttt{cond}, \texttt{case},
\texttt{let}, \texttt{and}, \texttt{or}, \texttt{fn} forms;
other forms of Clojure may be supported in the future but are not now.
In \texttt{let} bindings and \texttt{fn} argument lists,
Clojure's pattern matching mechanism for vector data type
(called vector destructuring) is supported. Also, compound
literals for vectors, hash maps, and
sets are supported just like in Clojure.

{\it TODO}

@@

\subsection{Tail Call}
\label{sec:tail}

Clojure provides special forms \texttt{loop} and \texttt{recur}
for writing tail-recursive programs. Such forms are not necessary
in Anglican, because Anglican programs are CPS-converted and do
not use the stack. For instance, no recursive call in Anglican
can lead to stack overflow.  In fact, in Anglican, it is
recommended to use recursive calls to functions instead of
\texttt{recur}. However, \texttt{loop}/\texttt{recur} is
provided in Anglican for convenience as a way to express loops.
\texttt{recur} outside of \texttt{loop} will lead to
unpredictable behaviour and hard-to-catch errors.

@@

\iftoggle{full}{\subsubsection }{\subsection}{Memoization}

>>>>

\documentclass[preprint]{sigplanconf}

@@

\usepackage{listings}
\usepackage{lmodern}
\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
\lstdefinestyle{default}{language=Anglican, basicstyle=\linespread{0.95}\ttfamily, columns=flexible, showstringspaces=false}

@@

An Anglican program, or \textit{query}, is compiled into a
Clojure function. When inference is performed with a provided
algorithm, this produces a sequence of samples. Anglican shares
a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican
can look like the following code:
%\begin{minipage}{\linewidth}
\begin{lstlisting}[style=default]
(defquery model "model selection" data
  (let [;;; Guess a distribution.
        dist (sample (categorical
                       [[normal 0.5]
                        [gamma 0.5]]))
        a (sample (gamma 1 1))
        b (sample (gamma 1 1))
        d (dist a b)]

@@

Anglican is a language for expressing probabilistic programs.
A probabilistic program is a stateful \textit{deterministic} computation
$\mathcal{P}$ with the following properties~\cite{TMP+15}:
\begin{itemize}
\item Initially, $\mathcal{P}$ expects no arguments.
\item On every call, $\mathcal{P}$ returns either a distribution
  and an address $(F,\alpha)$, a distribution and a value $(G,
        y)$, or a value $z$.
\item Upon returning $F$, $\mathcal{P}$ expects a value $x$ drawn from $F$ as
	the argument to the next call (\texttt{sample} in Anglican).
\item Upon returning $(G, y)$, $\mathcal{P}$ is invoked again
	without arguments (\texttt{observe} in Anglican).
\item Upon returning $z$, $\mathcal{P}$ terminates.
\end{itemize}
A program is run by calling $\mathcal{P}$ repeatedly until
termination.

A program need not generate the same sequence of random variables in every
execution. For this reason we assume that each random variable $x$ is assigned
a unique label $\alpha$, which we call an address, that induces a
correspondence between variables in different executions. Every execution
implicitly produces a sequence of triples of distributions, values of
\textit{latent} random variables, and addresses, $(F_i, x_i, \alpha_i)$.  We
call this sequence a \textit{trace} and denote it by $\pmb{x}$. A trace induces
a sequence of pairs $(G_j, y_j)$ of distributions and values of
\textit{observed} random variables. We call this sequence an \textit{image} and
denote it by $\pmb{y}$. For notational simplicity we assume that the program
always generates the same ordered set $\pmb{z}$ of \textit{results} $z_k$.

@@

The Anglican language is a subset of Clojure\footnote{It would be
possible to support almost full Clojure by expanding all macros
in the Anglican source code. However, in Clojure, unlike in
Scheme~\cite{SDF+10} or Common Lisp~\cite{PC94}, the result
of macro-expansion of derived special forms is not well
specified and implementation specific.}.  Anglican queries
are defined within \texttt{defquery}. The value of the last
expression in the query body is the result of the query.
\begin{lstlisting}[style=default]
(defquery name doc-string? param? expr*)
\end{lstlisting}
Anglican functions outside of a query are defined using
\texttt{defm} with the same syntax as Clojure \texttt{defn},
however \texttt{name} is bound to an Anglican function.
\begin{lstlisting}[style=default]
(defm name doc-string? [param*] expr*)
\end{lstlisting}
Within the body of \texttt{defquery} and \texttt{defm}, the
following special forms are supported, with syntax and semantics 
matching those of Clojure:
\begin{lstlisting}[style=default]
(if test then else?)
(when test expr*)
(cond clause*)
(case expr clause* default?)
(let [binding*] expr*)
(and expr*)
(or expr*)
(fn name? [param*] expr*)
(loop [binding*] expr*)
(recur expr*)
\end{lstlisting}
In \texttt{defquery} parameter, \texttt{let} bindings, and
\texttt{fn} argument lists, Clojure vector destructuring is
supported. Literals for atomic types and compound literals for
vectors, hash maps, and sets are supported just like in Clojure.

Clojure provides special forms \texttt{loop} and \texttt{recur}
for writing tail-recursive programs.  Anglican programs are
CPS-converted and do not use the stack; recursive calls
in Anglican cannot lead to stack overflow.  However,
\texttt{loop}/\texttt{recur} is provided in Anglican for
convenience as a way to express loops. Unlike in Clojure though,
\texttt{recur} supported only inside a \texttt{loop}.

@@

In addition to re-implementing a subset of Clojure, Anglican
provides special forms for probabilistic inference:
\begin{lstlisting}[style=default]
(sample address? distribution)
(observe address? distribution value)
(mem function)
(store tag value)
(retrieve tag)
\end{lstlisting}

\texttt{sample} and \texttt{observe} forms correspond to
handling $(F, \alpha)$ and $(G, y)$ correspondingly in the
definition of probabilistic program (Section~\ref{sec:pp-def}).
Their operational semantics can be defined in terms of
importance sampling(Equation~\ref{eqn:p-trace}): 
\begin{itemize}
	\item \texttt{sample} returns a value drawn from
		the distribution.
	\item \texttt{observe} multiplies the unnormalized
		probability of the program trace by the probability of
		drawing the value from the distribution.
\end{itemize}
However, the actual computation involved in handling of
\texttt{sample} and \texttt{observe} depends on the inference
algorithm. Different inference algorithms may treat
\texttt{sample} and \texttt{observe} differently as long as they
compute the same distribution of program results.
Denotational and operational semantics for an idealized version
of Anglican is introduced in \cite{SYH+16}; in general though,
specifying semantics of higher-order probabilistic programming
languages is an open problem.

\texttt{sample} and \texttt{observe} may appear anywhere in the
code of an Anglican program. To support these forms, the
inference engine must be able to intervene into execution of the
program, perform computations related to inference, and control
further execution of the program. This tight and complicated
interaction between the program and the inference engine
necessiates implementing Anglican as a language rather than a
library.

The remaining Anglican special forms support carrying values in
the hidden program state. \texttt{mem} implements stochastic
function  memoization (Section~\ref{sec:mem}). \texttt{store}
stores a value for a tag in the program state, which can be
later retrieved by the tag using \texttt{retrieve} in the same
program execution.

@@

\iftoggle{full}{\subsubsection }{\subsection}{Memoization}
\label{sec:mem}
